---
title: Every Tom, Dick, and Harry
author: Writing Functions in R
date: "R Pruim"
output: 
  slidy_presentation:
    widescreen: true
---

```{r setup, include = FALSE}
library(ggformula)
library(babynames)
library(dplyr)

knitr::opts_chunk$set(
  size = "small",
  fig.width = 6,
  fig.height = 2.5
)
```

## Loading some packages

```{r, message = FALSE}
library(babynames)
library(ggformula)
library(dplyr)
theme_set(theme_bw())
```

## What Ever Happened to Baby Jane (and Mary)?

Let's create a plot to see how the popularity of the names Jane and 
Mary has changed over time.

```{r}
babynames %>% 
  filter(name %in% c("Mary", "Jane"), sex == "F") %>%
  gf_line(prop ~ year, color = ~ name)
```



## What about Tom, Dick, and Harry?

We would like to generalize our code to handle any set of names
we want to compare.  In R we do this by writing **functions**.

We begin with our two questions:

 1. What do we want R to do?
   
    * This determines the name of the function (and what it does/returns)
    * `name_trend_plot()` might be a good name for the function
 
 2. What does it need to know to do this?
 
    * This determines the arguemnt list (list of inputs) to the function
    * The function will need to know which names to track.  
    * We should probably have an argument to restrict to just one sex.
    
## Creating a function in R

The general form of a fucntion definition is

```{r}
functionName <- 
  function( arg1 = default1, arg2 = default2, ... ) {
  # do stuff here
  # result of last line is returned.  
  # can also use explicit return().
}
```

## What about Tom, Dick, and Harry?

```{r}
name_trend_plot <- 
  function(names, gender) {
    babynames %>% 
      filter(name %in% names, sex == gender) %>%
      gf_line(prop ~ year, color = ~ name)
  }
```

## What about Tom, Dick, and Harry?

```{r}
name_trend_plot(c("Tom", "Dick", "Joe"), gender = "M")
```

## Adding more arguments

We can add arguments to control more features of the plot.

```{r}
name_trend_plot2 <- 
  function(names, gender = c("M", "F"), linewidth = 1) {
    babynames %>%
      filter(name %in% names, sex %in% gender) %>%
      gf_line(prop ~ year, color = ~ name, linetype = ~ sex, size = linewidth)
  }
```

```{r}
name_trend_plot2( c("Tom", "Dick", "Harry"), gender = "M", linewidth = 3)
```

```{r}
name_trend_plot2( c("Hillary", "Beverly"), linewidth = 1)
```

## Dot, Dot, Dot

R functions can have a special argument called `...`.  This argument
holds a list of any additional arguments. It is a very handy way to 
pass arguments along to another function.  (It has other uses, too.)

```{r}
name_trend_plot3 <- 
  function(names, gender = c("M", "F"), ...) {
    babynames %>%
      filter(name %in% names, sex %in% gender) %>%
      gf_line(prop ~ year, color = ~ name, linetype = ~ sex, ...)
  }
```

```{r}
name_trend_plot3(
  c("Robert", "James", "John"), gender = "M", size = 2,
  title = "Three Boys Names", caption = "Source: babynames")
```

